---
title: "Why grouping"
output: html_document
date: '2024-04-21'
---

In this document, the number of signatures for the regular NMF model is found. This is both used to argue for why the data should be grouped as well as for determining the number of signatures for the analysis.

Both BIC and cross validation is performed.

```{r setup, include=FALSE}
# To set a seed used for all. There is random initialization in the update functions
knitr::opts_chunk$set(seed = 123)
```

```{r}
# Libraries, working directory etc.
setwd("~/Desktop/Trio data/")
rm(list = ls())
library(tidyverse)
library(gridExtra)
library(ggplot2)
library(readr)
library(gtools)
library(dplyr)
library(lsa)
theme_set(theme_bw())
# Library for Hungarian method
library(clue)
```

```{r}
# Loading Poisson update function as well as other useful functions
source("~/Desktop/Trio data/Poisson updates function.R")
```

```{r}
# Loading the trio data
load("filtered_data_pat.RData")
load("filtered_data_mat.RData")

# Loading the matrices grouped by parents age with 40 groups
pat_data_matrix_40 <- readRDS("/Users/idajorgensen/Desktop/Trio data/CountMat_PaternalMutations_grouped_40.rds")
mat_data_matrix_40 <- readRDS("/Users/idajorgensen/Desktop/Trio data/CountMat_MaternalMutations_grouped_40.rds")

# Loading the matrices grouped by parents age with 119 groups
pat_data_matrix_119 <- readRDS("/Users/idajorgensen/Desktop/Trio data/CountMat_PaternalMutations_grouped_119.rds")
mat_data_matrix_119 <- readRDS("/Users/idajorgensen/Desktop/Trio data/CountMat_MaternalMutations_grouped_119.rds")

# Loading the full data matrices
full_pat_data_matrix <- readRDS("/Users/idajorgensen/Desktop/Trio data/CountMat_PaternalMutations.rds")
full_mat_data_matrix <- readRDS("/Users/idajorgensen/Desktop/Trio data/CountMat_MaternalMutations.rds")

# COSMIC signatures
COSMIC_v3.3.1_SBS_GRCh37 <- read.delim("~/Desktop/Trio data/COSMIC_v3.3.1_SBS_GRCh37.txt")
```

```{r}
# For choosing number of signatures
K <- c(2,3,4,5,6,7,8,9,10,11,12)
V_40 <- pat_data_matrix_40 
V_119 <- pat_data_matrix_119
Iter <- 10000
V_full <- full_pat_data_matrix
```


Here BIC values for both the data grouped in 40 groups, 119 groups and the full dataset are found.
```{r}
# BIC for Poisson, both groups
results_BIC_40 <- BIC_val(V_40,K,Iter, poisson_updates)
results_BIC_119 <- BIC_val(V_119,K,Iter, poisson_updates)
results_BIC_full <- BIC_val(V_full,K,Iter, poisson_updates)
BIC_values_40 <- results_BIC_40$BIC_values
BIC_values_119 <- results_BIC_119$BIC_values
BIC_values_full <- results_BIC_full$BIC_values

# Combine BIC values and corresponding K values into data frames
data_40 <- data.frame(K = K, BIC = BIC_values_40, Dataset = "40 groups")
data_119 <- data.frame(K = K, BIC = BIC_values_119, Dataset = "119 groups")
data_full <- data.frame(K = K, BIC = BIC_values_full, Dataset = "All groups")
combined_data <- rbind(data_40, data_119, data_full)

# Plot using ggplot2
ggplot(combined_data, aes(x = K, y = BIC, color = Dataset)) +
  geom_point() +
  labs(x = "Number of Signatures", y = "BIC Value", title = "BIC for Poisson Updates") +
  theme_minimal() +
  scale_color_manual(values = c("40 groups" = "blue", "119 groups" = "red", "All groups" = "green")) + 
  theme(axis.text = element_text(size = 14), axis.title = element_text(size = 14), legend.text = element_text(size = 14), legend.title = element_text(size = 14)) +
  scale_x_continuous(breaks = combined_data$K, labels = combined_data$K)
```

Plotting the BIC values for only the grouped datasets.
```{r}
# BIC plot with only grouped data

combined_data <- rbind(data_40, data_119)

# Plot using ggplot2
ggplot(combined_data, aes(x = K, y = BIC, color = Dataset)) +
  geom_point() +
  labs(x = "Number of Signatures", y = "BIC Value", title = "BIC for Poisson Updates") +
  theme_minimal() +
  scale_color_manual(values = c("40 groups" = "blue", "119 groups" = "red", "All groups" = "green")) + 
  theme(axis.text = element_text(size = 14), axis.title = element_text(size = 14), legend.text = element_text(size = 14), legend.title = element_text(size = 14)) +
  scale_x_continuous(breaks = combined_data$K, labels = combined_data$K)
```





Cross validation for both the 40 groups, the 119 groups and the full dataset
```{r}
# Define number of signatures with cross validation
J <- 20

results_cross_val_40 <- cross_val(V_40, K, Iter, J, poisson_updates)
Cost_40 <- results_cross_val_40$Cost_median
Cost_int_40 <- results_cross_val_40$Cost
results_cross_val_119 <- cross_val(V_119, K, Iter, J, poisson_updates)
Cost_119 <- results_cross_val_119$Cost_median
Cost_int_119 <- results_cross_val_119$Cost
results_cross_val_full <- cross_val(V_full, K, Iter, J, poisson_updates)
Cost_full <- results_cross_val_full$Cost_median
Cost_int_full <- results_cross_val_full$Cost


# Calculate mean and standard deviation of cost values for each K
mean_cost_40 <- apply(Cost_int_40, 2, mean)
sd_cost_40 <- apply(Cost_int_40, 2, sd)

mean_cost_119 <- apply(Cost_int_119, 2, mean)
sd_cost_119 <- apply(Cost_int_119, 2, sd)

mean_cost_full <- apply(Cost_int_full, 2, mean)
sd_cost_full <- apply(Cost_int_full, 2, sd)

# Combine mean and standard deviation into data frames
confidence_data_40 <- data.frame(K = K, Mean_Cost = mean_cost_40, SD_Cost = sd_cost_40, Dataset = "40 groups")
confidence_data_119 <- data.frame(K = K, Mean_Cost = mean_cost_119, SD_Cost = sd_cost_119, Dataset = "119 groups")
confidence_data_full <- data.frame(K = K, Mean_Cost = mean_cost_full, SD_Cost = sd_cost_full, Dataset = "All groups")

# Combine cross validation values and corresponding K values into data frames
data_40 <- data.frame(K = K, Cost = Cost_40, Dataset = "40 groups")
data_119 <- data.frame(K = K, Cost = Cost_119, Dataset = "119 groups")
data_full <- data.frame(K = K, Cost = Cost_full, Dataset = "All groups")
combined_data <- rbind(data_40, data_119, data_full)

# Plot using ggplot2
ggplot(combined_data, aes(x = K, y = Cost, color = Dataset)) +
  geom_point() +
  geom_line(data = combined_data, aes(x = K, y = Cost, color = Dataset, group = Dataset)) + 
  geom_ribbon(data = confidence_data_40, aes(x = K, y = Mean_Cost, ymin = Mean_Cost - SD_Cost, ymax = Mean_Cost + SD_Cost, fill = Dataset), alpha = 0.3) +
  geom_ribbon(data = confidence_data_119, aes(x = K, y = Mean_Cost, ymin = Mean_Cost - SD_Cost, ymax = Mean_Cost + SD_Cost, fill = Dataset), alpha = 0.3) + 
  geom_ribbon(data = confidence_data_full, aes(x = K, y = Mean_Cost, ymin = Mean_Cost - SD_Cost, ymax = Mean_Cost + SD_Cost, fill = Dataset), alpha = 0.3) +
  labs(x = "Number of Signatures", y = "Cross validation Value", title = "Crossvalidation for Poisson Updates") +
  theme_minimal() +
  scale_color_manual(values = c("40 groups" = "blue", "119 groups" = "red", "All groups" = "green")) + 
  scale_fill_manual(values = c("40 groups" = "lightblue", "119 groups" = "lightcoral", "All groups" = "lightgreen")) + 
  theme(axis.text = element_text(size = 14), axis.title = element_text(size = 14), , legend.text = element_text(size = 14), legend.title = element_text(size = 14)) +
  scale_x_continuous(breaks = combined_data$K, labels = combined_data$K)
```

Plot of the cross validation for only the grouped datsets.
```{r}
# Plot with only the grouped data
combined_data <- rbind(data_40, data_119)
ggplot(combined_data, aes(x = K, y = Cost, color = Dataset)) +
  geom_point() +
  geom_line(data = combined_data, aes(x = K, y = Cost, color = Dataset, group = Dataset)) + 
  geom_ribbon(data = confidence_data_40, aes(x = K, y = Mean_Cost, ymin = Mean_Cost - SD_Cost, ymax = Mean_Cost + SD_Cost, fill = Dataset), alpha = 0.3) +
  geom_ribbon(data = confidence_data_119, aes(x = K, y = Mean_Cost, ymin = Mean_Cost - SD_Cost, ymax = Mean_Cost + SD_Cost, fill = Dataset), alpha = 0.3) + 
  labs(x = "Number of Signatures", y = "Cross validation Value", title = "Crossvalidation for Poisson Updates") +
  theme_minimal() +
  scale_color_manual(values = c("40 groups" = "blue", "119 groups" = "red", "All groups" = "green")) + 
  scale_fill_manual(values = c("40 groups" = "lightblue", "119 groups" = "lightcoral", "All groups" = "lightgreen")) + 
  theme(axis.text = element_text(size = 14), axis.title = element_text(size = 14), , legend.text = element_text(size = 14), legend.title = element_text(size = 14)) +
  scale_x_continuous(breaks = combined_data$K, labels = combined_data$K)
```
















