---
title: "Why grouping"
output: html_document
date: '2024-04-21'
---

In this document, the number of signatures for the regular NMF model is found. This is both used to argue for why the data should be grouped as well as for determining the number of signatures for the analysis.

Both BIC and cross validation is performed.

```{r setup, include=FALSE}
# To set a seed used for all. There is random initialization in the update functions
knitr::opts_chunk$set(seed = 123)
set.seed(NULL)
knitr::opts_chunk$set(seed = NULL)
```

```{r}
# Libraries, working directory etc.
setwd("~/Desktop/Trio data/")
rm(list = ls())
library(tidyverse)
library(gridExtra)
library(ggplot2)
library(readr)
library(gtools)
library(dplyr)
library(lsa)
theme_set(theme_bw())
# Library for Hungarian method
library(clue)
```

```{r}
# Loading Poisson update function as well as other useful functions
source("~/Desktop/Trio data/Poisson updates function.R")
```

```{r}
# Loading the trio data
load("filtered_data_pat.RData")
load("filtered_data_mat.RData")

# Loading the matrices grouped by parents age with groups
pat_data_matrix_grouped <- readRDS("/Users/idajorgensen/Desktop/Trio data/CountMat_PaternalMutations_grouped.rds")
mat_data_matrix_grouped <- readRDS("/Users/idajorgensen/Desktop/Trio data/CountMat_MaternalMutations_grouped.rds")

# Loading the full data matrices
full_pat_data_matrix <- readRDS("/Users/idajorgensen/Desktop/Trio data/CountMat_PaternalMutations.rds")
full_mat_data_matrix <- readRDS("/Users/idajorgensen/Desktop/Trio data/CountMat_MaternalMutations.rds")

# COSMIC signatures
COSMIC_v3.3.1_SBS_GRCh37 <- read.delim("~/Desktop/Trio data/COSMIC_v3.3.1_SBS_GRCh37.txt")
```

```{r}
# For choosing number of signatures
K <- c(2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20)
V_grouped <- pat_data_matrix_grouped
V_grouped_mat <- mat_data_matrix_grouped
Iter <- 30000
V_full <- full_pat_data_matrix
```


Here BIC values for both the grouped data and the full dataset are found.
```{r, fig.width= 5, fig.height=3.5}
par(mai = c(0.5,0.5,0.3,0.05), mfrow = c(2,1), cex = 1, mgp = c(1.5, 0.5, 0))
# BIC for both groups and full data
results_BIC_grouped <- BIC_val(V_grouped,K,Iter, poisson_updates)
results_BIC_full <- BIC_val(V_full,K,Iter, poisson_updates)
BIC_values_grouped <- results_BIC_grouped$BIC_values
BIC_values_full <- results_BIC_full$BIC_values

# Combine BIC values and corresponding K values into data frames
data_grouped <- data.frame(K = K, BIC = BIC_values_grouped, Dataset = "Grouped data")
data_full <- data.frame(K = K, BIC = BIC_values_full, Dataset = "Full data")
combined_data <- rbind(data_grouped, data_full)

# Plot using ggplot2
ggplot(combined_data, aes(x = K, y = BIC, color = Dataset)) +
  geom_point() +
  labs(x = "Number of Signatures", y = "BIC Value") +
  theme_minimal() +
  scale_color_manual(values = c("Grouped data" = "blue", "Full data" = "forestgreen")) + 
  theme(axis.text = element_text(size = 18), axis.title = element_text(size = 18), legend.text = element_text(size = 18), legend.title = element_text(size = 18)) +
  scale_x_continuous(breaks = combined_data$K, labels = combined_data$K)
```

Plotting the BIC values for only the grouped dataset.
```{r, fig.width= 10, fig.height=3.5}
par(mai = c(0.5,0.5,0.3,0.05), mfrow = c(2,1), cex = 1, mgp = c(1.5, 0.5, 0))
# BIC plot with only grouped data

combined_data <- rbind(data_grouped)

# Plot using ggplot2
ggplot(combined_data, aes(x = K, y = BIC, color = Dataset)) +
  geom_point() +
  labs(x = "Number of Signatures", y = "BIC Value") +
  theme_minimal() +
  scale_color_manual(values = c("Grouped data" = "blue", "119 groups" = "red", "All groups" = "green")) + 
  theme(axis.text = element_text(size = 18), axis.title = element_text(size = 18), legend.text = element_text(size = 18), legend.title = element_text(size = 18), legend.position = "none") +
  scale_x_continuous(breaks = combined_data$K, labels = combined_data$K)
```





Cross validation for both the grouped data and the full dataset
```{r, fig.width= 5, fig.height=3.5}
par(mai = c(0.5,0.5,0.3,0.05), mfrow = c(2,1), cex = 1, mgp = c(1.5, 0.5, 0))
# Define number of signatures with cross validation
J <- 20

results_cross_val_grouped <- cross_val(V_grouped, K, Iter, J, poisson_updates)
Cost_grouped <- results_cross_val_grouped$Cost_median
Cost_int_grouped <- results_cross_val_grouped$Cost
# saveRDS(list(Cost_matrix = Cost_int_grouped, Cost_medians = Cost_grouped), file = "paternal_normal_nmf_cross_val.RDS")
results_cross_val_full <- cross_val(V_full, K, Iter, J, poisson_updates)
Cost_full <- results_cross_val_full$Cost_median
Cost_int_full <- results_cross_val_full$Cost

# results_cross_val_grouped <- readRDS("paternal_normal_nmf_cross_val.RDS")
Cost_grouped <- results_cross_val_full$Cost_median
Cost_int_grouped <- results_cross_val_full$Cost_matrix

# Calculate mean and standard deviation of cost values for each K
mean_cost_grouped <- apply(Cost_int_grouped, 2, mean)
sd_cost_grouped <- apply(Cost_int_grouped, 2, sd)

mean_cost_full <- apply(Cost_int_full, 2, mean)
sd_cost_full <- apply(Cost_int_full, 2, sd)

# Combine mean and standard deviation into data frames
confidence_data_grouped <- data.frame(K = K, Mean_Cost = mean_cost_grouped, SD_Cost = sd_cost_grouped, Dataset = "Grouped data")
confidence_data_full <- data.frame(K = K, Mean_Cost = mean_cost_full, SD_Cost = sd_cost_full, Dataset = "Full data")

# Combine cross validation values and corresponding K values into data frames
data_grouped <- data.frame(K = K, Cost = Cost_grouped, Dataset = "Grouped data")
data_full <- data.frame(K = K, Cost = Cost_full, Dataset = "Full data")
combined_data <- rbind(data_grouped, data_full)

# Plot using ggplot2
ggplot(combined_data, aes(x = K, y = Cost, color = Dataset)) +
  geom_point() +
  geom_line(data = combined_data, aes(x = K, y = Cost, color = Dataset, group = Dataset)) + 
  geom_ribbon(data = confidence_data_grouped, aes(x = K, y = Mean_Cost, ymin = Mean_Cost - SD_Cost, ymax = Mean_Cost + SD_Cost, fill = Dataset), alpha = 0.3) +
  geom_ribbon(data = confidence_data_full, aes(x = K, y = Mean_Cost, ymin = Mean_Cost - SD_Cost, ymax = Mean_Cost + SD_Cost, fill = Dataset), alpha = 0.3) +
  labs(x = "Number of Signatures", y = "Cross validation Value") +
  theme_minimal() +
  scale_color_manual(values = c("Grouped data" = "blue", "Full data" = "forestgreen")) + 
  scale_fill_manual(values = c("Grouped data" = "lightblue", "Full data" = "lightgreen")) + 
  theme(axis.text = element_text(size = 18), axis.title = element_text(size = 18), , legend.text = element_text(size = 18), legend.title = element_text(size = 18)) +
  scale_x_continuous(breaks = combined_data$K, labels = combined_data$K)
```



Plotting only cross validation for the grouped data
```{r, fig.width= 10, fig.height=3.5}
par(mai = c(0.5,0.5,0.3,0.05), mfrow = c(2,1), cex = 1, mgp = c(1.5, 0.5, 0))
# Plot with only the grouped data
combined_data <- rbind(data_grouped)
ggplot(combined_data, aes(x = K, y = Cost, color = Dataset)) +
  geom_point() +
  geom_line(data = combined_data, aes(x = K, y = Cost, color = Dataset, group = Dataset)) + 
  geom_ribbon(data = confidence_data_grouped, aes(x = K, y = Mean_Cost, ymin = Mean_Cost - SD_Cost, ymax = Mean_Cost + SD_Cost, fill = Dataset), alpha = 0.3) +
  labs(x = "Number of Signatures", y = "Kullback-Leibler divergence") +
  theme_minimal() +
  scale_color_manual(values = c("Grouped data" = "blue", "119 groups" = "red", "All groups" = "green")) + 
  scale_fill_manual(values = c("Grouped data" = "lightblue", "119 groups" = "lightcoral", "All groups" = "lightgreen")) + 
  theme(axis.text = element_text(size = 18), axis.title = element_text(size = 18), , legend.text = element_text(size = 18), legend.title = element_text(size = 18), legend.position = "none") +
  scale_x_continuous(breaks = combined_data$K, labels = combined_data$K)
```














### For the maternal!!

BIC values for grouped dataset for maternal
```{r, fig.width= 10, fig.height=3.5}
par(mai = c(0.5,0.5,0.3,0.05), mfrow = c(2,1), cex = 1, mgp = c(1.5, 0.5, 0))
# BIC for grouped data
results_BIC_grouped <- BIC_val(V_grouped_mat,K,Iter, poisson_updates)
BIC_values_grouped <- results_BIC_grouped$BIC_values

# Combine BIC values and corresponding K values into data frames
data_grouped <- data.frame(K = K, BIC = BIC_values_grouped, Dataset = "Grouped data")
combined_data <- rbind(data_grouped)

# Plot using ggplot2
ggplot(combined_data, aes(x = K, y = BIC, color = Dataset)) +
  geom_point() +
  labs(x = "Number of Signatures", y = "BIC Value") +
  theme_minimal() +
  scale_color_manual(values = c("Grouped data" = "blue", "Full data" = "forestgreen")) + 
  theme(axis.text = element_text(size = 18), axis.title = element_text(size = 18), legend.text = element_text(size = 18), legend.title = element_text(size = 18), legend.position = "none") +
  scale_x_continuous(breaks = combined_data$K, labels = combined_data$K)
```




Cross validation for maternal grouped data
```{r, fig.width= 10, fig.height=3.5}
par(mai = c(0.5,0.5,0.3,0.05), mfrow = c(2,1), cex = 1, mgp = c(1.5, 0.5, 0))
# Define number of signatures with cross validation
J <- 20

results_cross_val_grouped <- cross_val(V_grouped_mat, K, Iter, J, poisson_updates)
Cost_grouped <- results_cross_val_grouped$Cost_median
Cost_int_grouped <- results_cross_val_grouped$Cost
# saveRDS(list(Cost_matrix = Cost_int_grouped, Cost_medians = Cost_grouped, K = K, BIC_values_grouped = BIC_values_grouped), file = "maternal_normal_nmf_model_selection_results.RDS")

# results_cross_val_grouped <- readRDS("maternal_normal_nmf_model_selection_results.RDS")
# Cost_grouped <- results_cross_val_grouped$Cost_median
# Cost_int_grouped <- results_cross_val_grouped$Cost_matrix

# Calculate mean and standard deviation of cost values for each K
mean_cost_grouped <- apply(Cost_int_grouped, 2, mean)
sd_cost_grouped <- apply(Cost_int_grouped, 2, sd)

# Combine mean and standard deviation into data frames
confidence_data_grouped <- data.frame(K = K, Mean_Cost = mean_cost_grouped, SD_Cost = sd_cost_grouped, Dataset = "Grouped data")

# Combine cross validation values and corresponding K values into data frames
data_grouped <- data.frame(K = K, Cost = Cost_grouped, Dataset = "Grouped data")
combined_data <- rbind(data_grouped)

# Plot using ggplot2
ggplot(combined_data, aes(x = K, y = Cost, color = Dataset)) +
  geom_point() +
  geom_line(data = combined_data, aes(x = K, y = Cost, color = Dataset, group = Dataset)) + 
  geom_ribbon(data = confidence_data_grouped, aes(x = K, y = Mean_Cost, ymin = Mean_Cost - SD_Cost, ymax = Mean_Cost + SD_Cost, fill = Dataset), alpha = 0.3) +
  labs(x = "Number of Signatures", y = "Kullback-Leibler divergence") +
  theme_minimal() +
  scale_color_manual(values = c("Grouped data" = "blue", "Full data" = "forestgreen")) + 
  scale_fill_manual(values = c("Grouped data" = "lightblue", "Full data" = "lightgreen")) + 
  theme(axis.text = element_text(size = 18), axis.title = element_text(size = 18), , legend.text = element_text(size = 18), legend.title = element_text(size = 18), legend.position = "none") +
  scale_x_continuous(breaks = combined_data$K, labels = combined_data$K)
```


















#### Test with BIC only non-zero entries

```{r, fig.width= 5, fig.height=3.5}
par(mai = c(0.5,0.5,0.3,0.05), mfrow = c(2,1), cex = 1, mgp = c(1.5, 0.5, 0))
# BIC for Poisson, both groups
results_BIC_grouped <- BIC_val_new(V_grouped_mat,K,Iter, poisson_updates)
BIC_values_grouped <- results_BIC_grouped$BIC_values


# Combine BIC values and corresponding K values into data frames
data_grouped <- data.frame(K = K, BIC = BIC_values_grouped, Dataset = "Grouped data")
combined_data <- rbind(data_grouped)

# Plot using ggplot2
ggplot(combined_data, aes(x = K, y = BIC, color = Dataset)) +
  geom_point() +
  labs(x = "Number of Signatures", y = "BIC Value") +
  theme_minimal() +
  scale_color_manual(values = c("Grouped data" = "blue", "Full data" = "forestgreen")) + 
  theme(axis.text = element_text(size = 18), axis.title = element_text(size = 18), legend.text = element_text(size = 18), legend.title = element_text(size = 18)) +
  scale_x_continuous(breaks = combined_data$K, labels = combined_data$K)
```

